\section{Library {\tt H}: hygienic macros}

\paragraph{Warning} This hygienic macro system is quite new and
experimental. Its API is likely to evolve over the next versions of
metalua. Feedbacks are especially welcome.

A common problem with meta-programming tools is variable capture, and
the art of automatically avoiding capture is called hygiene. the {\tt
  H} library automates as much as possible the handling of hygiene.

The design of H tries to respect metalua's core principles:
\begin{itemize}
\item Nothing taboo under the hood: the underlying mechanisms of the
  language must remain simple enough to be intelligible to their
  intended users. Black magic should be banned from the desing. This
  rules out hygienic macros as a primitive: these either rely on very
  advanced and hard to predict mechanisms, or severely limit the
  manipulation tools available to the macro authors.
\item Simple by default: advanced users should know what happens under
  the hood, but more casual users should be able to simply turn the
  ignition and drive. It should be possible to use H, for regular
  macros, without much understanding of its advanced principles and
  implementation.
\item Everything's a regular program: again, most macro systems
  offering hygiene limit macro manipulations to a term rewriting
  framework, which might be Turing complete, but makes many advanced
  stuff cumbersome to write. AST are regular data, which must be
  manipulable by regular programs.
\item Extension: metalua tries to offer the most extensible possible
  language to its users. If the author of the language couldn't
  implement a feature as a regular extension, it would probably
  outline a severe limitation of the system.
\end{itemize}

\paragraph{Inside captures}
There are two kind of captures, inside a macro and outside a
macro. People often think about inside captures, in parts because the
C preprocessor is subject to it. It happens when a macro inserts user
code in a quote, and the quote declares a local variable that shadows
a user one:

\begin{verbatim}
-{ block:
   require 'dollar'
   function dollar.TRICOND(cond, iftrue, iffalse)
      local code = +{ block:
         local tmp, result
         tmp = -{cond}
         if tmp then result = -{iftrue} 
         else result = -{iffalse} end }
      return `Stat{ code, +{result} }
   end }

local tmp = 5
local foo = $TRICOND(tmp%2==0, "even", "odd")
\end{verbatim}

Here, the \verb|tmp| local variable used in the macro code captures
the user's one, and cause a failure (an attempt to get the modulus of
\verb|nil|). The expanded code looks like:
\begin{Verbatim}
local tmp = 5
local foo = -{ `Stat{ +{ block:
   local tmp, result
   tmp = tmp%2==0
   if tmp then result = "even"
   else result = "odd" end
}, +{result} } }
\end{Verbatim}

This is fixed by renaming automatically all local variables in the
macro with fresh names. H provides an AST walker which does
that. However, it needs to rename only macro code, not user-provided
code; therefore the macro writer has to somehow mark the user
code. This is done with the ``!'' prefix operator: sub-trees marked
with ``!'' won't experience any renaming. The following version is
therefore safe w.r.t. inside variable capture:

\begin{verbatim}
-{ block:
   require 'dollar'
   function dollar.TRICOND(cond, iftrue, iffalse)
      local code = +{ block:
         local tmp, result
         tmp = -{!cond}
         if tmp then result = -{!iftrue} 
         else result = -{!iffalse} end }
      return H(`Stat{ code, +{result} })
   end }

local tmp = 5
local foo = $TRICOND(tmp%2==0, "even", "odd")
\end{verbatim}

It expands to:

\begin{Verbatim}
local tmp = 5
local foo = -{ `Stat{ +{ block:
   local -{`Id '.1.L.tmp'}, -{`Id '.2.L.result'} -- new fresh names
   -{`Id '.1.L.tmp'} = tmp%2==0 -- no capture!
   if -{`Id '.1.L.tmp'} then -{`Id '.2.L.result'} = "even"
   else -{`Id '.2.L.result'} = "odd" end
}, `Id '.2.L.result' } }
\end{Verbatim}

\paragraph{Outside captures}
We've seen that macros can capture the user's variables; but the
opposite can also happen: the user can capture the variables required
by a macro:

\begin{Verbatim}
-{ block:
   require 'dollar'
   dollar.log = |x| +{ printf("%s = %s", 
                              -{table.tostring(x)}, 
                              table.tostring(-{x})) } }
local x = { 1, 2, 3 }
\$log(x) -- prints "`Id 'x' = { 1, 2, 3 }"
\end{Verbatim}

This expands to:
\begin{Verbatim}
printf("%s = %s", "`Id 'x'", table.tostring(x))
\end{Verbatim}

Now, replace "x" with "table", and you get an outside capture: "local
table = { 1, 2, 3 } shadows the table module from which the macro
tries to get table.tostring().

The most widespread language which supports non-hygienic macros,
Common Lisp, deals with that issue by being a Lisp-2: it has separate
namespaces for functions and ``normal'' variables. This happens to
remove many common capture cases. 

H fixes this by renaming the free variables in hygienized
macros. After this, a "local new\_names = old\_names" statement is
generated, which re-establishes the correspondance between
names. Let's make the examples above hygienic:

\begin{Verbatim}
-{ block:
   require 'dollar'
   dollar.log = |x| H+{ printf("%s = %s", 
                               -{!table.tostring(x)}, 
                               table.tostring(-{!x})) } }
local table = { 1, 2, 3 }
\$log(table) -- prints "`Id 'table' = { 1, 2, 3 }"
\end{Verbatim}

The code above expands into:

\begin{Verbatim}
local table = { 1, 2, 3 }
(-{`Id '.1.X.printf'}) ("%s = %s",
                        "`Id 'table'",
                        (-{`Id '.2.X.table'}.tostring(table)))
\end{Verbatim}

To make this work, we need to introduce, somewhere where no variable
is captured, the following local statement:
\begin{Verbatim}
local -{`Id '.1.X.printf'}, -{`Id '.2.X.table'} = printf, table
\end{Verbatim}

The correct way would be to let the user decide where this statement
should go.

