\section{Extension {\tt match}: structural pattern matching}
Pattern matching is an extremely pleasant and powerful way to handle tree-like
structures, such as ASTs. Unsurprisingly, it's a feature found in most
ML-inspired languages, which excel at compilation-related tasks. There is a
pattern matching extension for metalua, which is extremely useful for most
meta-programming purposes.

\subsection{Purpose}

First, to clear a common misconception: structural pattern matching has
absolutely nothing to do with regular expresssions matching on strings: it works
on arbitrary structured data.

When manipulating trees, you want to check whether they have a certain structure
(e.g. a `Local{ } node with as first child a list of variables whose tags are
`Id{ }); when you've found a data that matches a certain pattern, you want to
name the interesting sub-parts of it, so that you can manipulate them easily;
finally, most of the time, you have a series of different possible patterns, and
you want to apply the one that matches a given data. These are the needs
addressed by pattern matching: it lets you give a list of ({\tt pattern ->
  code\_to\_execute\_if\_match}) associations, selects the first matching
pattern, and executes the corresponding code. Patterns both describe the
expected structures and bind local variables to interesting parts of the data.
Those variables' scope is obviously the code to execute upon matching success.

\paragraph{Match statement} 
A match statement has the form:

\begin{verbatim}
match <some_value> with
| <pattern_1> -> <block_1>
| <pattern_2> -> <block_2>
...
| <pattern_n> -> <block_n>
end
\end{verbatim}

The first vertical bar after the "with" is optional; moreover, a pattern can
actually be a list of patterns, separated by bars. In this case, it's enough for
one of them to match, to get the block to be executed:

\begin{verbatim}
match <some_value> with
| <pattern_1> | <pattern_1_bis > -> <block_1>
...
end
\end{verbatim}

When the match statement is executed, the first pattern which matches
{\tt<some\_value>} is selected, the corresponding block is executed, and all
other patterns and blocks are ignored. If no pattern matches, an error
{\tt"mismatch"} is raised. However, we'll see it's easy to add a catch-all
pattern at the end of the match, when we want it to be failproof.

\subsection{Patterns definition}

\paragraph{Atomic litterals}
Syntactically, a pattern is mostly identical to the values it matches: numbers,
booleans and strings, when used as patterns, match identical values.

\begin{verbatim}
match x with
| 1 -> print 'x is one'
| 2 -> print 'x is two'
end
\end{verbatim}

\paragraph{Tables}
Tables as patterns match tables with the same number of array-part elements, if
each pattern field matches the corresponding value field. For instance, \{1, 2,
3\} as a pattern matches \{1, 2, 3\} as a value. Pattern \{1, 2, 3\} matches
value \{1, 2, 3, foo=4\}, but pattern \{1, 2, 3, foo=4\} doesn't match value
\{1, 2, 3\}: there can be extra hash-part fields in the value, not in the
pattern. Notice that field 'tag' is a regular hash-part field, therefore \{1, 2,
3\} matches `Foo\{1, 2, 3\} (but not the other way around). Of course, table
patterns can be nested. The table keys must currently be integers or strings.
It's not difficult to add more, but the need hasn't yet emerged.

If you want to match tables of arbitrary array-part size, you can add a "..." as
the pattern's final element. For instance, pattern \{1, 2, ...\} will match all
table with at least two array-part elements whose two first elements are 1 and
2.

\paragraph{Identifiers}
The other fundamental kind of patterns are identifiers: they match everything,
and bind whatever they match to themselves. For instance, pattern {1, 2, x} will
match value {1, 2, 3}, and in the corresponding block, local variable x will be
set to 3. By mixing tables and identifiers, we can already do interesting
things, such as getting the identifiers list out of a local statement, as
mentionned above:

\begin{verbatim}
match stat with
| `Local{ identifiers, values } ->
   table.foreach(identifiers, |x| print(x[1])
... -- other cases
end
\end{verbatim}

When a variable appears several times in a single pattern, all the elements they
match must be equal, in the sense of the "==" operator. Fore instance, pattern \{
  x, x \} will match value \{ 1, 1 \}, but not \{ 1, 2 \}. Both values would be
matched by pattern \{ x, y \}, though. A special identifier is "\_", which doesn't
bind its content. Even if it appears more than once in the pattern, metched
value parts aren't required to be equal. The pattern "\_" is therefore the
simplest catch-all one, and a match statement with a "{\tt| \_ ->}" final
statement will never throw a "mismatch" error.

\paragraph{Guards}

Some tests can't be performed by pattern matching. For these cases, the pattern
can be followed by an "if" keyword, followed by a condition.

\begin{verbatim}
match x with
| n if n%2 == 0 -> print 'odd'
| _ -> print 'even'
end
\end{verbatim}

Notice that the identifiers bound by the pattern are available in the guard
condition. Moreover, the guard can apply to several patterns:

\begin{verbatim}
match x with
| n | {n} if n%2 == 0 -> print 'odd'
| _ -> print 'even'
end
\end{verbatim}

\paragraph{Multi-match}

If you want to match several values, let's say 'a' and 'b', there's an easy way:

\begin{verbatim}
match {a,b} with
| {pattern_for_a, pattern_for_b} -> block
...
end
\end{verbatim}

However, it introduces quite a lot of useless tables and checks. Since this kind
of multiple matches are fairly common, they are supported natively:

\begin{verbatim}
match a, b with
| pattern_for_a, pattern_for_b -> block
...
end
\end{verbatim}

This will save some useless tests and computation, and the compiler will
complain if the number of patterns doesn't match the number of values.

\paragraph{String regular expressions}
There is a way to use Lua's regular exressions with match, through the division
operator ``/'': the left operand is expected to be a literal string, interpreted
as a regular expression. The variables it captures are stored in a table, which
is matched as a value against the right-hand-side operand. For instance, the
following case succeeds when {\tt foo} is a string composed of 3 words separated
by spaces. In case of success, these words are bound to variables {\tt w1}, {\tt
  w2} and {\tt w3} in the executed block:

\begin{verbatim}
match foo with
| "^(%w+) +(%w+) +(%w+)$"/{ w1, w2, w3 } -> 
   do_stuff (w1, w2, w3)
end
\end{verbatim}

\subsection{Examples}
There are quite a lot of samples using match in the metalua distribution, and
more will come in the next one. Dig in the samples for fairly simple usages, or
in the standard libs for more advanced ones. Look for instance at examples
provided with the {\tt walk} library.
