\def\tableHeader{\begin{tabular}{|c|c|p{5cm}|}\hline
\bf name & \bf type & \multicolumn{1}{c|}{\bf description} \\\hline}
\def\entry#1#2#3{{#1} & {\tt#2} & {#3} \\\hline}
\def\tableFooter{\hline\end{tabular}}

\section{{\tt mlp}, the metalua parser}

Metalua parser is built on top of \verb|gg|, and cannot be understood
without some knowledge of it. Basically, \verb|gg| allows not only to
build parsers, but to build {\em extensible} parsers. Depending on a
parser's type (sequence, sequence set, list, expression\ldots),
different extension methods are available, which are documented in
\verb|gg| reference. The current section will give the information
needed to extend Metalua syntax:
\begin{itemize}
\item what \verb|mlp| entries are accessible for extension;
\item what do they parse;
\item what is the underlying parser type (and therefore, what
  extension methods are supported)
\end{itemize}

\vfill\pagebreak

\subsection{Parsing expressions}
\tableHeader

\entry{mlp.expr}{gg.expr}{Top-level expression parser, and the main
  extension point for Metalua expression. Supports all of the methods
  defined by {\tt gg.expr}.}

\entry{mlp.func\_val}{gg.sequence}{Read a function definition,
  from the arguments' openning parenthesis to the final {\tt end}, but
  excluding the initial {\tt function} keyword, so that it can be used
  both for anonymous functions, for {\tt function some\_name(...) end}
  and for {\tt local function some\_name(...) end}.}

% \entry{mlp.func\_params\_content}{gg.list}{Read a potentially empty
%   (``{\tt)}''- or ``{\tt|}''-terminated) list of function definition
%   parameters, i.e. identifiers or ``{\tt ...}'' varargs. Surrounding
%   parentheses are excluded. Don't get confused between parameters and
%   arguments: parameters are the variable names used in a function
%   definition; arguments are the values passed in a function call.}

% \entry{mlp.func\_args\_content}{gg.list}{Read a potentially emtpy list
%   of function call arguments. Surrounding parentheses are excluded.}

% \entry{mlp.func\_args}{gg.sequence\_set}{Read function arguments: a
%   list of expressions between parenthses, or a litteral table, or a
%   litteral string.}

%\entry{mlp.func\_params}{}{}
\entry{mlp.expr\_list}{}{}

%\entry{mlp.adt}{\rm custom function}{Read an algebraic datatype
%  without its leading backquote.}

\entry{mlp.table\_content}{gg.list}{Read the content of a table,
  excluding the surrounding braces}

\entry{mlp.table}{gg.sequence}{Read  a litteral table,
  including the surrounding braces}

\entry{mlp.table\_field}{\rm custom function}{Read a table entry: {\tt
    [foo]=bar}, {\tt foo=bar} or {\tt bar}.}

\entry{mlp.opt\_id}{\rm custom function}{Try to read an identifier, or
  an identifier splice. On failure, returns false.}

\entry{mlp.id}{\rm custom function}{Read an identifier, or
  an identifier splice. Cause an error if there is no identifier.}

\tableFooter

\vfill\pagebreak

\subsection{Parsing statements}
\tableHeader
\entry{mlp.block}{gg.list}{Read a sequence of statements, optionally
  separated by semicolons. When introducing syntax extensions, it's
  often necessary to add block terminators with {\tt
  mlp.block.terminators:add().}}
\entry{mlp.for\_header}{\rm custom function}{Read a {\tt for} header,
from just after the ``{\tt for}'' to just before the ``{\tt do}''.}
\entry{mlp.stat}{gg.multisequence}{Read a single statement.}
\tableFooter

Actually, {\tt mlp.stat} is an extended version of a multisequence: it
supports easy addition of new assignment operator. It has a field {\tt
assignments}, whose keys are assignment keywords, and values are
assignment builders taking left-hand-side and right-hand-side as
parameters. for instance, C's ``+='' operator could be added as:
\begin{verbatim}
mlp.lexer:add "+="
mlp.stat.assignments["+="] = function (lhs, rhs)
  assert(#lhs==1 and #rhs==1)
  local a, b = lhs[1], rhs[1]
  return +{stat: (-{a}) = -{a} + -{b} }
end 
\end{verbatim}

\subsection{Other useful functions and variables}

\begin{itemize}
\item{\tt mlp.gensym()} generates a unique identifier. The uniqueness
  is guaranteed, therefore this identifier cannot capture another
  variable; it is useful to write hygienic\footnote{Hygienic macros
    are macros which take care not to use names that might interfere
    with user-provided names. The typical non-hygienic macro in C
    is {\tt \#define SWAP( a, b) \{ int c=a; a=b; b=c; \}}: this macro
    will misearbly fail if you ever call it with a parameter named
    {\tt c}. There are well-known techniques to automatically make a
    macro hygienic. Without them, you'd have to generate a unique name
    for the temporary variable, if you had a {\tt gensym()} operator
    in C's preprocessor} macros. 
\end{itemize}
