\subsection{Exceptions}
As a first non-trivial example of extension, we'll pick exception:
there is a mechanism in Lua, {\tt pcall()}, which essentially provides
the raw functionality to catch errors when some code is run, so
enhancing it to get full exceptions is not very difficult. We will aim
at:
\begin{itemize}
\item Having a proper syntax, the kind you get in most
  exception-enabled languages;
\item being able to easily classify exception hierarchically;
\item being able to attach additional data to exception (e.g. an error
  message);
\item not interfere with the usual error mechanism;
\item support the ``finally'' feature, which guaranties that a piece
  of code (most often about resource liberation) will be executed.
\end{itemize}

\subsubsection{Syntax}
There are many variants of syntaxes for exceptions. I'll pick one
inspired from OCaml, which you're welcome to dislike. And in case you
dislike it, writing one which suits your taste is an excellent
exercice. So, the syntax for exceptions will be something like:
\begin{Verbatim}[fontsize=\scriptsize]

try
   <protected block of statements>
with
  <exception_1> -> <block of statements handling exception 1>
| <exception_2> -> <block of statements handling exception 2>
  ...
| <exception_n> -> <block of statements handling exception n>
end
\end{Verbatim}

Notice that OCaml lets you put an optional ``{\tt|}'' in front of the
first exception case, just to make the whole list look more regular,
and we'll accept it as well. Let's write a {\tt gg} grammar parsing
this:

\begin{Verbatim}[fontsize=\scriptsize]

trywith_parser = 
   gg.sequence{ "try",  mlp.block,  "with",  gg.optkeyword "|",
                gg.list{ gg.sequence{ mlp.expr,  "->",  mlp.block },
                         separators = "|", terminators = "end" },
                "end", 
                builder = trywith_builder }
mlp.stat:add(trywith_parser)
mlp.lexer:add{ "try", "with", "->" }
mlp.block.terminator:add{ "|", "with" }
\end{Verbatim}

We use {\tt gg.sequence} to chain the various parsers; {\tt
  gg.optkeyword} lets us allow the optional ``{\tt|}''; {\tt gg.list}
lets us read an undetermined series of exception cases, separated by
keyword ``{\tt|}'', until we find the terminator ``{\tt end}''
keyword. The parser delegates the building of the resulting statement
to {\tt trywith\_builder}, which will be detailled later. Finally, we
have to declare a couple of mundane things:
\begin{itemize}
\item that {\tt try}, {\tt with} and {\tt->} are keywords. If we don't
  do this, the two firsts will be returned by the lexer as identifiers
  instead of keywords; the later will be read as two separate keywords
  ``{\tt-}'' and ``{\tt>}''. We don't have to declare explicitly
  ``{\tt|}'', as single-character symbols are automatically considered to
  be keywords.
\item that ``{\tt|}'' and ``{\tt with}'' can terminate a block of
  statements. Indeed, metalua needs to know when it reached the end of
  a block, and introducing new constructions which embed blocks often
  introduce new block terminators. In our case, ``{\tt with}'' marks
  the end of the block in which exceptions are monitored, and ``{\tt|}''
  marks the beginning of a new exception handling case, and therefore
  the end of the previous case's block.
\end{itemize}

That's it for syntax, at least for now. The next step is to decide
what kind of code we will generate.

The fundamental mechanism is {\tt pcall(func, arg1, arg2, ...,
  argn)}: this call will evaluate\\
{\tt func(arg1, arg2, ..., argn)}, and:
\begin{itemize}
\item if everything goes smoothly, return {\tt true}, followed by any
  value(s) returned by {\tt func()};
\item if an error occurs, return {\tt false}, and the error object,
  most often a string describing the error encountered.
\end{itemize}

We'll exploit this mechanism, by enclosing the guarded code in a
function, calling it inside a {\tt pcall()}, and using special error
objects to represent exceptions.

\subsubsection{Exception objects}
We want to be able to classify exceptions hierarchically: each
exception will inherit form a more generic exception, the most generic
one being simply called ``{\tt exception}''. We'll therefore design a
system which allows to specialize an exception into a sub-exception,
and to compare two exceptions, to know whether one is a special case
of another. Comparison will be handled by the usual {\tt< > <= >=}
operators, which we'll overload through metatables. Here is an
implementation of the base exception {\tt exception}, with working
comparisons, and a {\tt new()} method which allow to specialize an
exception. Three exceptions are derived as an example, so that
{\tt exception > exn\_invalid > exn\_nullarg} and {\tt exception >
  exn\_nomorecoffee}:

\begin{Verbatim}[fontsize=\scriptsize]

exception = { } ; exn_mt = { } 
setmetatable (exception, exn_mt)

exn_mt.__le = |a,b| a==b or a<b 
exn_mt.__lt = |a,b| getmetatable(a)==exn_mt and 
                    getmetatable(b)==exn_mt and
                    b.super and a<=b.super

function exception:new()
   local e = { super = self, new = self.new }
   setmetatable(e, getmetatable(self))
   return e
end

exn_invalid     = exception:new()
exn_nullarg     = exn_invalid:new()
exn_nomorecofee = exception:new()
\end{Verbatim}

To compile a {\tt try/with} block, after having put the guarded block
into a {\tt pcall()} we need to check whether an exception was raised,
and if is has been raised, compare it with each case until we find one
that fits. If none is found (either it's an uncaught exception, or a
genuine error which is not an exception at all), it must be rethrown. 

Notice that throwing an exception simply consists into sending it as
an error:
\begin{Verbatim}[fontsize=\scriptsize]

throw = error
\end{Verbatim}

To fix the picture, here is some simple code using {\tt try/catch},
followed by its translation:

\begin{Verbatim}[fontsize=\scriptsize]

-- Original code:
try
   print(1)
   print(2)
   throw(exn_invalid:new("toto"))
   print("You shouldn't see that")
with
| exn_nomorecofee -> print "you shouldn't see that: uncomparable exn"
| exn_nullarg     -> print "you shouldn't see that: too specialized exn"
| exn_invalid     -> print "exception caught correctly"
| exception       -> print "execution should never reach that far"
end 
print("done")
\end{Verbatim}

\begin{Verbatim}[fontsize=\scriptsize]

-- Translated version:
local status, exn = pcall (function ()
   print(1)
   print(2)
   throw(exn_invalid)
   print("You shouldn't see that")
   end)

if not status then
   if exn < exn_nomorecoffee then
      print "you shouldn't see that: uncomparable exn"
   elseif exn < exn_nullarg then
      print "you shouldn't see that: too specialized exn"
   elseif exn < exn_invalid then
      print "exception caught correctly"
   elseif exn < exception then
      print "execution should never reach that far"
   else error(exn) end
end 
print("done")
\end{Verbatim}

In this, the only nontrivial part is the sequence of {\tt
  if/then/elseif} tests at the end. If you check the doc about AST
representation of such blocks, you'll come up with some generation
code which looks like:

\pagebreak

\begin{Verbatim}[fontsize=\scriptsize]

function trywith_builder(x)
   ---------------------------------------------------------
   -- Get the parts of the sequence:
   ---------------------------------------------------------
   local block, _, handlers = unpack(x)

   ---------------------------------------------------------
   -- [catchers] is the big [if] statement which handles errors
   -- reported by [pcall].
   ---------------------------------------------------------
   local catchers = `If{ }
   for _, x in ipairs (handlers) do
      -- insert the condition:
      table.insert (catchers, +{ -{x[1]} <= exn })
      -- insert the corresponding block to execute on success:
      table.insert (catchers, x[2])
   end

   ---------------------------------------------------------
   -- Finally, put an [else] block to rethrow uncought errors:
   ---------------------------------------------------------
   table.insert (catchers, +{error (exn)})

   ---------------------------------------------------------
   -- Splice the pieces together and return the result:
   ---------------------------------------------------------
   return +{ block:
      local status, exn  = { pcall (function() -{block} end) }
      if not status then
         -{ catchers }
      end }
end
\end{Verbatim}

\subsubsection{Not getting lost between metalevels}
This is the first non-trivial example we see, and it might require a
bit of attention in order not to be lost between metalevels. Parts of
this library must go at metalevel (i.e. modify the parser itself at
compile time), other parts must be included as regular code:
\begin{itemize}
\item {\tt trywith\_parser} and {\tt trywith\_builder} are at metalevel:
  they have to be put between \verb|-{...}|, or to be put in a file
  which is loaded through \verb|-{ require ... }|. 
\item the definitions of {\tt throw}, the root {\tt exception} and the
  various derived exceptions are regular code, and must be included normally.
\end{itemize}

The whole result in a single file would therefore look like:

\begin{Verbatim}[fontsize=\scriptsize]

-{ block:
   local trywith_builder = ...
   local trywith_parser  = ...
   mlp.stat:add ...
   mlp.lexer:add ...
   mlp.block.terminator:add ... }

throw     = ...
exception = ...
exn_mt    = ...

exn_invalid     = ...
exn_nullarg     = ...
exn_nomorecofee = ...

-- Test code
try
   ...
with
| ... -> ...
end
\end{Verbatim}

Better yet, it should be organized into two files:
\begin{itemize}
\item the parser modifier, i.e. the content of ``{\tt-\{block:...\}}''
  above, goes into a file ``ext-syntax/exn.lua'' of Lua's path;
\item the library part, i.e. {\tt throw ... exn\_nomorecoffee ...}
  goes into a file ``ext-lib/exn.lua'' of Lua's path;
\item the sample calls them both with metalua standard lib's {\tt
extension} function:
\end{itemize}

\begin{Verbatim}[fontsize=\scriptsize]

-{ extension "exn" }
try
  ...
with
| ... -> ...
ene
\end{Verbatim}

\subsubsection{shortcomings}
This first attempt is full of bugs, shortcomings and other
traps. Among others:
\begin{itemize}
\item Variables {\tt exn} and {\tt status} are subject to capture;
\item There is no way to put personalized data in an exception. Or,
  more accurately, there's no practiccal way to retrieve it in the
  exception handler.
\item What happens if there's a {\tt return} statement in the guraded
  block?
\item There's no {\tt finally} block in the construction.
\item Coroutines can't yield across a {\tt pcall()}. Therefore, a
yield in the guarded code will cause an error.
\end{itemize}

Refining the example to address these shortcomings is left as an
exercice to the reader, we'll just give a couple of design
hints. However, a more comprehensive implementation of this exception
system is provided in metalua's standard libraries; you can consider
its sources as a solution to this exercice!

\subsubsection{Hints}
Addressing the variable capture issue is straightforward: use {\tt
  mlp.gensym()} to generate unique identifiers (which cannot capture
anything), and put anti-quotes at the appropriate places. Eventually,
metalua will include an hygienization library which will automate this
dull process. 

Passing parameters to exceptions can be done by adding arbitrary
parameters to the {\tt new()} method: these parameters will be stored
in the exception, e.g. in its array part. finally, the
syntax has to be extended so that the caught exception can be given a
name. Code such as the one which follows should be accepted:

\begin{Verbatim}[fontsize=\scriptsize]
try
   ...
   throw (exn_invalid:new "I'm sorry Dave, I'm afraid I can't do that.")
with
| exn_invalid e -> printf ("The computer choked: %s", e[1])
end
\end{Verbatim}

The simplest way to detect user-caused returns is to create a unique
object (typically an empty table), and return it at the end of the
block. when no exception has been thrown, test whether that object was
returned: if anything else than it was returned, then propagate it (by
{\tt return}ing it again). If not, do nothing. Think about the case
when multiple values have been returned.

The {\tt finally} block poses no special problem: just go through it,
whether an exception occured or not. Think also about going through it
even if there's a {\tt return} to propagate.

As for yielding from within the guarded code, there is a solution,
which you can find by searching Lua's mailing list archives. The idea
is to run the guarded code inside a coroutine, and check what's
returned by the coroutine run:
\begin{itemize} 
\item if it's an error, treat it as a {\tt pcall()} returning false;
\item if it's a normal termination, treat it as a {\tt pcall()}
  returning true;
\item if it's a yield, propagate it to the upper level; when resumed,
  propagate the resume to the guarded code which yielded.
\end{itemize}