\def\function#1{\paragraph{#1}}

\section{Standard library}
Metalua comes with a standard library which extends Lua's one. These
extended features are enabled by adding a ``{\tt require "std"}''
statement in your source files (it is already required at the
compile-time level, so there's no need for a 
``{\tt -\{ require "std" \}}''). 

\subsection{Base library extensions}

\function{min(...)}
Return the min of its arguments, according to {\tt<}. There must be at
least one argument.

\function{max(...)}
Return the max of its arguments, according to {\tt<}. There must be at
least one argument.

\function{o(...)}  
Return the composition of all functions passed as arguments. For
instance, {\tt printf} could be defined as {\tt print `o`
  string.format}\footnote{Or in regular Lua syntax {\tt o(print,
    string.format)}.}

\function{id(...)}
Return all its arguments unchanged.

\function{const (...)}
Return a function which always returns {\tt ...}, whatever its arguments.
For instance, {\tt const(1, 2, 3)(4, 5, 6)} returns {\tt1, 2, 3}.

\function{printf(fmt,...)}
Equivalent to {\tt print(string.format(fmt,...))}.

\function{values(table)}
Iterator, to be used in a {\tt for} loop, which returns all the values
of {\tt table}. 

{\tt for x in values(t) do [...] end} is equivalent to {\tt for \_, x in
  pairs(t) do [...] end}.

\function{keys(table)}
Iterators, to be used in a {\tt for} loop, which return all the keys
of {\tt table}. 

{\tt for k in keys(t) do [...] end} is equivalent to {\tt for k, \_ in
  pairs(t) do [...] end}.

\function{extension(name)}
FIXME: move this into metalua.compiler?

\subsection{{\tt table} extensions}

Many of the extensions of {\tt table} are dedicated to a more
functional style programming. When compared to the functions in
Haskell or ML's standard libs, these one are slightly more general,
taking advantage from Lua's dynamic typing.

\function{table.iforeach(f, ...)}

{\tt table.iforeach(f, t)} will evaluate f with every array-part
elements of t in order.

If more than one table are passed as parameters, f will receive an
element of each table at each iteration. For instance, {\tt
  table.iforeach (print, \{1, 2, 3\}, \{4, 5, 6\}, \{7, 8, 9\})} will
print:
\begin{verbatim}
1 4 7
2 5 8
3 6 9
\end{verbatim}

If the second and/or third parameters are numbers, they indicate the
first and last indexes to use in the tables. First index defaults to
1, last index default to the length of the longest table. If only one
number is passed, it's considered to be the first index. For instance,
{\tt table.iforeach (print, 2, \{1, 2, 3\}, \{4, 5, 6\}, \{7, 8, 9\})}
will only print:
\begin{verbatim}
2 5 8
3 6 9
\end{verbatim}

\function{table.imap(f, ...)}

Similar to {\tt table.iforeach()}, except that the results of {\tt
  f()} calls are collected into a list and returned.

For instance, {\tt table.imap((|x,y|x+y), 2, \{"foo", 1, 2, 3\},
  \{"bar", 10, 20, 30\})} will return {\tt\{11, 22, 33\}}.

\function{table.ifold(f, acc, ...)}  

Fold list elements thanks to a combining function {\tt f()}, which
takes two list elements and returns one result. For the first
iteration, {\tt f()} takes {\tt acc} as its first param. For instance,
the sum of {\tt list}'s elements can be computed by {\tt table.ifold(
  (|x,y| x+y), 0, list)}.

This function also accepts first and last indexes after {\tt acc}, and
more than one table argument: if there are more than one table, then
more than two parameters are passed to {\tt f()}. For instance, this
function returns $\sum_{i\le2} \max(x[i], y[i])$: {\tt
  table.ifold( (|acc, xi, yi| acc + max (xi, yi)), 0, 2, x, y)}.

\function{table.izip(...)}  

Take a sequence of lists, and return the
list of their first elements, then their second elements, etc. For
instance, {\tt table.izip (\{1,2,3\}, \{4,5,6\})} will return
{\tt\{\{1,4\}, \{2,5\} , \{3,6\}\}}.

\function{table.ifilter(f, t)}

Return the list of all elements of {\tt t} for which {\tt f} returns
neither {\tt nil} nor {\tt false}.

\function{table.icat(...)}

Concatenate all the lists passed as arguments into a single one, then
return it.

\function{table.iflatten(x)}

Flatten a list of lists into a list. for instance, {\tt
  table.iflatten\{\{1,2\}, \{3,4\}\}} returns {\tt\{1,2,3,4\}}.

\function{table.irev(t)}
Reverse the order of elements in {\tt t}'s array-part. This is done
in-place: if you don't want to alter the original list, first copy
it.

\function{table.iall(f, ...)}

Return true if and only if {\tt table.iforeach(f, ...)} would return only
non-false values.

\function{table.iany(f, ...)}

Return true if and only if {\tt table.iforeach(f, ...)} would return
at least one non-false value.

\function{table.shallow\_copy(t)}

Does a shallow copy of the table {\tt t}. This differs from {\tt
  table.icat(t)}, because the latter would only copy tha array-part,
whereas the former also copies the hash-part.

\function{table.deep\_copy(t)}

Does a deep copy of {\tt t}, i.e. all keys and values are recursively
copied. Handles tables with shared and circular references correctly;
also set the copy's metatable to the original's one.

\function{table.range(a, b, c)}

Return a list of all integers between {\tt a} and {\tt b} inclusive,
with an increment {\tt c} (which defaults to 1).

\function{table.tostring(t, ...)}

Return a string which represents {\tt t}. This string is correctly
indented, and handles Metalua's special syntax for ADT/AST
gracefully. If {\tt"nohash"} is passed as an additional argument, then
only the tag and array-part of the table are displayed. If a number
{\tt n} is passed as extra argument, the function tries to keep the
number of characters per line under {\tt n}.

\function{table.print(t, ...)}

Equivalent to {\tt print(table.tostring(t, ...))}.

\subsection{{\tt string} extensions}

\function{string.split(string, pattern)}
Cut {\tt string} into a list of substrings separated by pattern {\tt
  pattern}, and return that list.


\function{string.strmatch(...)}

Alias for {\tt string.match}: since it's quite common in metalua to
use the pattern matching extension, which turns {\tt match} into a
keyword, it's practical to have another name for this function.

\subsection{Library {\tt mlc}}

FIXME: move in metalua.compiler.

This library offer conversion between the different possible
representations of metalua programs:
\begin{itemize}
\item as source files
\item as compiled files
\item as source strings
\item as compiled chunk dumps (which are actually strings)
\item as lexeme streams
\item as AST
\item FIXME
\end{itemize}

Hopefully, the function names are self-explanatory. Some of them are
simply aliases to other standard functions such as {\tt loadstring()}
or {\tt string.dump()}; many others are compositions of other
functions. The point is that every sensible transformation from
representation {\tt xxx} to representation {\tt yyy} should appear in
this library under the name {\tt yyy\_of\_xxx()}. This way, users
don't have to wonder how to chain the appropriate functions to get the
expected result.

The functions available in this module are:

\begin{tabular}{|l|l|l|l|}
  \hline
  FIXME \\\hline
\end{tabular}

FIXME: implementation sucks beyond maintainability, it should be
rewritten.

\subsection{Library {\tt walker}}
This library allows code-walking, i.e. applying advanced, non-local 
transformations on ASTs. It's powerful, but definitely not user
friendly; eventuallty, it might be replaced by a Term Rewriting
System, supported by its own Domain-Specific Language.

\function{walk\_stat (cfg)}
FIXME
%don't get confused between the AST it applies on, the the AST being
%currently inspected.

%function calls and method invocations in a statement context are not
%considered as expressions.

\function{walk\_expr (cfg)}
Same as {\tt walk\_stat}, except that it takes an expression AST
instead of a statement AST. 

\function{walk\_block (cfg)}
Same as {\tt walk\_stat}, except that it takes a statements block AST
instead of a single statement AST.
