\section[Metalua extensions]{Metalua syntax extensions over Lua}
Metalua is essentially Lua + code generation at compile time +
extensible syntax. However, there are a couple of additional
constructs, considered of general interest, which have been added to
Lua's original syntax. These are presented in this section

\subsection{Anonymous functions}
Lua lets you use anonymous functions. However, when programming in a
functional style, where there are a lot of short anonymous functions
simply returning an expression, the default syntax becomes
cumbersome. Metalua being functional-styel friendly, it offers a
terser idiom: ``{\tt function(arg1, arg2, argn) return some\_expr
  end}'' can be written:\\
``{\tt|arg1,arg2,argn| some\_exp}''.

Notice that this notation is currying-friendly, i.e. one can easily
write functions that return functions: ``{\tt function(x) return
function(y) return x+y end end}'' is simply written ``{\tt|x||y|
x+y}''.

Lua functions can return several values, but it appeared that
supporting multiple return values in metalua's short lambda notation
caused more harm than good. If you need multiple returns, use the
traditional long syntax.

Finally, it's perfectly legal to define a parameterless function, as
in {\tt | | 42}. This makes a convenient way to pass values around in a
lazy way.

\subsection{Functions as infix operators}

In many cases, people would like to extend syntax simply to create
infix binary operators. Haskell offers a nice compromize to satisfy
this need without causing any mess, and metalua incorporated it: when
a function is put between backquotes, it becomes infix. for instance,
let's consider the {\tt plus} function ``{\tt plus=|x,y|x+y}''; this
function can be called the classic way, as in ``{\tt plus (20, 22)}''; but
if you want to use it in an infix context, you can also write ``{\tt
20 `plus` 22}''.

\subsection{Algebraic datataypes}

This syntax for datatypes is of special importance to metalua, as it's
used to represent source code being manipulated. Therefore, it has its
dedicated section later in this manual.

\subsection{Metalevel shifters}

These to dual notations are the core of metaprogramming: one
transforms code into a manipulaeble representation, and the other
transforms the representation back into code. They are noted
{\tt+\{...\}} and {\tt-\{...\}}, and due to their central role in
metalua, their use can't be summed up adequately here: they are fully
described in the subsequent sections about metaprogramming.
