\section{Extension {\tt trywith}: exceptions and finalization}
Lua offers error handling primitives \verb+pcall()+ and
\veb+xpcall()+. However, they are pretty low level, and their syntax
is cumbersome to use and read back. This extension offers a proper
syntax for the handling of such exceptions.

\subsection{Syntax}
An error handling statement has the following form:

\begin{verbatim}
try
   <protected block>
catch <exception pattern #1> then
   <exception handling block #1>
catch <exception pattern #2> then
   <exception handling block #2>
   ...
catch <exception pattern #n> then
   <exception handling block #n>
finally
   <finalization block>
end
\end{verbatim}

\subsection{Semantics}
When such a statement is executed:
\begin{itemize}
\item the protected code is executed
\item if its execution causes an error, the error message is matched
  against all exception patterns, until one matches or the last one is
  reached (see the \verb+match+ extension for patterns
  semantics). Patterns can include guard clauses. The block
  corresponding to the first matching pattern is executed. If no
  pattern matches, the error will be rethrown.
\item the block following the \verb+finally+ keyword will be executed
  no matter what: if the protected block executes succesfully, if it
  raises an error, if it causes a \verb+return+, if an error occurs in
  an exception handling block, whether an error is caught or
  not... The only reason why the finalization block might not be run
  is because of a sudden death of the process (call to {\tt os.exit()}
  or core dump).
\end{itemize}

The finally block can be omitted, or there can be no error catching
case. The following examples are legal:

\begin{verbatim}
try
   f = io.open ('file.txt', 'r')
   num_char = #f:read '*a'
finally
   f:close()
end

try
   do_stuff()
catch "mismatch" then
   print "match statement failure"
catch "[Ee]of"/_ then -- regexp pattern
   print "An end-of-file error seems to have happened"
catch x if type(x)=='table' then
   print "The error is a table, not a string!"
end
\end{verbatim}

 \subsection{RAII}
 RAII, or ``Resource Acquisition Is Initialization'', is a common
 programming pattern in which an object's lifetime is strictly
 associated with a given lexical scope. For instance, if a file is
 opened in a given scope, it must be closed as soon as this scope is
 leaved, even if it's leaved due to a {\tt return} or an error. The
 ``finally'' block allows this, but since it's a very common use case,
 there is a dedicated extension ``with/do'': you initialize some resource
 behind the ``with'' keyword, and it will be closed after the ``do''
 block is left. The only constraint is that the resources must have a
 {\tt:close()} method which releases them. Here is a usage example:

\begin{verbatim}
-{ extension 'withdo' }

with f1, f2 = io.open 'file1.txt', io.open 'file2.txt' do
   local t1 = f1:read '*a'
   local t2 = f2:read '*a'
   printf("The files contain %i and %i chars respectively", t1, t2)
end
\end{verbatim}

